---
title: "Microbiome analysis using phyloseq"
---
This R Markdown contains a starting collection of basic microbiome analysis using pyhloseq (https://github.com/joey711/phyloseq). A tutorial can be accessed here: https://joey711.github.io/phyloseq/import-data.html

###1. Set up - install packages and load functions
If you have not yet installed thi packages => first install (usig https://github.com/janinazuend/Analysis-16S-rRNA/blob/main/setup.Rmd)
```{r setup, include=FALSE}
mylib = 'C:\\Users\\zuendj\\Documents\\R4Libs'
.libPaths(mylib)

# devtools::install_github("tidyverse/tidyverse") 

library(phyloseq)
library(tidyverse)
library(openxlsx)
library(microViz)
library(rstatix)
library(ggpubr)
library(plyr)
library(ggpmisc)
library(cowplot)
library(vegan)


```

```{r}
#download the functions and store them locally on your computer=> from fconstancias: https://zenodo.org/records/6473394 
#load them into R => replace directory with the directory on your computer
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_taxa_tests.R") 
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_normalisation.R") 
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_alpha.R") 
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_beta.R") 
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_heatmap.R")
source("C:/Users/zuendj/Desktop/03_data R/divcom_functions/R/phyloseq_taxa_tests.R")

#function for getting metadata file
sample.data.frame <- function(ps) {
  return(as(phyloseq::sample_data(ps), "data.frame"))}

#function to name ggplot facets
facet_labeller <- function(variable,value){
  return(names[value])
}
```

### 1. Load/create a phyloseq object 
For different methods to create phyloseq see: https://joey711.github.io/phyloseq/import-data.html

```{r}
ps<-readRDS("C:/Users/zuendj/Desktop/03_data R/fibreGum/phyloseq_FibreGum_Run1.rds")
```

### 2. Check the Library size & Rarefaction (Normalization => for Diversity analysis !!!)
```{r}
# Extract the OTU (or ASV) table as a matrix
otu_mat <- as(otu_table(ps), "matrix")

# If taxa are rows, transpose to have samples in rows
if(taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

# Calculate library size per sample (sum of counts per sample)
library_sizes <- setNames(as.data.frame(rowSums(otu_mat)), c("Libsize"))%>%
  arrange(Libsize) ##order
```
```{r}
#plot the read depth
vegan::rarecurve(otu_mat, step=10000, cex=0.5)
```
#### Rarefying the reads
```{r}
ps %>%
  rarefy_even_depth(rngseed = 123,
                    sample.size = 1095326
                    ) -> ps_rare
```
##  2. Alpha diversity
```{r}
#alpha diversity
ps_rare%>%
  phyloseq_alphas(phylo = F) -> alphas
```

```{r}
alphas%>%
  mutate(Arm = factor(Arm, levels =c("Intervention", "Ctrl")))%>%
  gather(index, alpha_value, "observed":"diversity_shannon")%>%
  mutate(index = factor(index, levels= c("observed", "diversity_shannon")))%>%
  ggplot(aes(x = Arm,
             y = alpha_value)) +
  facet_grid(scale="free", space="free_x", rows=vars(index)) +
  geom_boxplot(aes(group=Time), outlier.shape = NA) +
  geom_point(aes(group=Time, colour=Time), position= position_jitterdodge(0.3), alpha=0.75,   size = 3)+
  ylab("Diversity") + 
  xlab(NULL) +
  theme(panel.background = element_rect(fill="white"),
        panel.border=element_rect(fill=NA))+
  ggtitle(title)->alpha_plot

show(alpha)

ggsave(filename = "alpha_c.jpeg",
       path ="C:/Users/zuendj/Desktop/03_data R/wp1/output/final",
       width = 17,
       height = 17,
       units = c("cm"))
```

```{r}

alphas%>%
  gather(index, alpha_value, "observed":"diversity_shannon")%>%
  mutate(index = factor(index, levels= c("observed", "diversity_shannon")))%>%
  group_by(index, condition)%>%
  shapiro_test(alpha_value)

#shannon is not normal for 3C and BHI like => should be wilcox test but conditions are anyway not significant 
alphas%>%
   mutate(condition = factor(condition, levels =c("feces", "7C",   "3C+Muc", "7C-Muc" , "3C" ,   "Glc"  ,"H2O"
                                                 ,"BHI like", "GMM like"
                                                 )))%>%
  gather(index, alpha_value, "observed":"diversity_shannon")%>%
  mutate(index = factor(index, levels= c("observed", "diversity_shannon")))%>%
  group_by(index)%>%
  t_test(alpha_value ~ condition)%>%
  add_xy_position(x= "condition", fun="max") %>% 
  add_significance()%>%
  mutate(y.position =ifelse(index=="observed" & group1=="feces" & group2=="Glc", 160, y.position))%>%
  mutate(y.position =ifelse(index=="observed" & group1=="feces" & group2=="3C", 140, y.position))%>%
  mutate(y.position =ifelse(index=="diversity_shannon" & group1=="feces" & group2=="Glc", 4.5, y.position))-> stat.test


title<-paste("Alpha diversity")
names<-list("observed"= "Observed", "diversity_shannon"="Shannon")

alphas%>%
  mutate(condition = factor(condition, levels =c("feces", "7C",   "3C+Muc", "7C-Muc" , "3C" ,   "Glc"  ,"H2O"
                                                 ,"BHI like", "GMM like"
                                                 )))%>%
  gather(index, alpha_value, "observed":"diversity_shannon")%>%
  mutate(index = factor(index, levels= c("observed", "diversity_shannon")))%>%
  ggplot(aes(x = condition,
             y = alpha_value)) +
  facet_grid(scale="free", space="free_x", rows=vars(index), labeller=facet_labeller) +
  geom_boxplot(aes(group=condition), outlier.shape = NA) +
  geom_point(aes(colour =donor_name, group=condition), position= position_jitterdodge(0.3), alpha=0.75,   size = 3)+
  ylab("Diversity") + 
  xlab(NULL) +
  geom_point(data = data.frame(alpha_value=5, condition="feces", index ="diversity_shannon"), colour="white")+
  geom_point(data = data.frame(alpha_value=175, condition="feces", index ="observed"), colour="white")+
  theme(axis.text.y= element_text(size = 20), 
            axis.title.y = element_text(size=20), 
            axis.title.x = element_text(size=20),
            axis.text.x = element_text(size=20, angle=90, vjust=0.2, hjust=1),
            legend.text = element_text(size = 20),
            legend.title = element_text(size= 20),
            plot.title = element_text(size=20),
          strip.text.y = element_text(size=20), 
        panel.background = element_rect(fill="white"),
        panel.border=element_rect(fill=NA), axis.ticks.x = element_blank())+
  ggtitle(title)+
  scale_x_discrete(labels = c('feces' = 'Feces', 
                              '7C' = '6C+Muc', 
                              '3C+Muc' = '3C+Muc', 
                              "Nut"="Nut", 
                              "SS"="SS",
                              "7C-Muc"="6C", 
                              "3C"="3C", 
                              "H2O"="C-depleted", 
                              "Glc"="Glucose", 
                              "BHI like" = "BHI-like", 
                              "GMM like"= "GMM-like"))+
  stat_pvalue_manual(stat.test, hide.ns = T, label = "p.adj.signif", size=5, label.size = 8)+
  labs(colour="Donor")->alpha

show(alpha)

ggsave(filename = "alpha_c.jpeg",
       path ="C:/Users/zuendj/Desktop/03_data R/wp1/output/final",
       width = 17,
       height = 17,
       units = c("cm"))
```

```{r}
alphas%>%
  subset(!condition %in% "feces")->feces
median(feces$diversity_shannon)
aggregate(alphas$observed, by=list(c(alphas$condition)), FUN=median)
```

## Compositional bar plots for each donor
```{r}
#fix tax table
noNA = !is.na(tax_table(phyloseq_rare)[,"Genus"]) & !is.na(tax_table(phyloseq_rare)[,"Species"])
tax_table(phyloseq_rare)[noNA][,"Species"] = paste(tax_table(phyloseq_rare)[noNA][,"Genus"], tax_table(phyloseq_rare)[noNA][,"Species"])
phyloseq_rare%>%
  tax_fix()->phyloseq_rare
```

```{r}
meta<-sample.data.frame(phyloseq_rare)
#define order for bar plot
meta%>%
  mutate(condition = factor(condition, levels=c("feces", "7C",   "3C+Muc", "7C-Muc" , "3C" ,   "Glc"  ,"H2O" 
                                                 ,"BHI like", "GMM like" 
                                                )))->sample_data(phyloseq_rare)
```

```{r}
setwd(mydir)
myPal <- tax_palette(
  data = phyloseq_rare, rank = "Family", n = 10, pal = "greenArmytage",
  add = c(Other = "white")
)

myPal["Lachnospiraceae"] <- "lightblue"
myPal["Bacteroidaceae"] <- "steelblue"
myPal["Enterobacteriaceae"] <- "yellowgreen"
myPal["Prevotellaceae"] <- "forestgreen"
myPal["Ruminococcaceae"] <- "lightpink1"
myPal["Bifidobacteriaceae"] <- "red3"
myPal["Oscillospiraceae"] <- "sienna1"
myPal["Streptococcaceae"] <- "orange"
myPal["Sutterellaceae"] <- "khaki1"
myPal["Erysipelatoclostridiaceae"] <- "orchid4"
myPal["Other"] <- "lightgrey"
tax_palette_plot(myPal)

phyloseq_rare%>%
  ps_arrange(condition)%>%
  microViz::comp_barplot(
    tax_level = "Family",
    label = "condition",
    n_taxa = 10,
    bar_width = 0.9,
    sample_order = "default",
    tax_transform_for_plot = "compositional",
    palette= myPal) +
    ylab("Proportion") + xlab("")-> p_hist

p_hist + facet_grid(cols=vars(donor_name), scales = "free", space = "free_x", drop = TRUE)+
  theme(text= element_text(size=20),
        axis.text.y = element_text(size=20),
        axis.text.x = element_text(size=20, angle=90, vjust=0.2, hjust=1, margin=margin(-8,0,0,0)),
        axis.ticks.x = element_blank(),
        legend.text = element_text(size=20, vjust=1, hjust=0),
        panel.spacing = unit(0.05, "cm"),
        strip.text = element_text(size=20), plot.title = element_text(size=20)) + 
  scale_y_continuous(breaks=c(0.25, 0.5, 0.75))+ 
  ggtitle("Taxonomic distribution and ASV representation in cultures and feces")-> p_hist

show(p_hist)

ggsave(filename = "Barplots_family_noleg.png",
       path = mydir,
       width = 40,
       height = 15,
       units = c("cm"))
```
## Distance plots of pairwise comparisons
```{r}
phyloseq_rare %>%
  phyloseq::distance(method = "bray") -> bc

phyloseq_rare %>%
  ordinate(method = "PCoA",
           distance = bc) -> ord
```

```{r}
setwd(mydir)
phyloseq_rare %>% 
  phyloseq_distance_boxplot(p = ., # phyloseq object
                            dist = bc, # distance matrix
                            d = "condition") -> out # column in metadata for comparaisons
#get the distances
setNames(as.data.frame(out$matrix),c("sample_name", "sample_name_comp", "value", "condition_comp", "comparison")) -> distances

left_join(left_join(distances%>%
  subset(comparison %in% c("feces")), meta),
                    setNames(aggregate(distances$value, by=list(distances$condition_comp), FUN=median), c("condition_comp", "mean_value")))%>%
  mutate(sample_type = factor(sample_type, levels= c("feces", "culture")))->joined_2

###get fecal IDs
meta[, c(1,3)]%>%
  subset(sample_name %in% c("M-81","M-82", "M-83", "M-84", "M-85", "M-86", "M-87", "M-88"))%>%
  setNames(c("sample_name_comp", "donor_name_comp"))->donor_comp

joined_3<-left_join(joined_2, donor_comp)

#filter out for comparisons between feces and respective donor microbiota
for (i in 1:nrow(joined_3)){
  if (joined_3[i,c("donor_name")] == joined_3[i,c("donor_name_comp")]){joined_3[i,c("match")]= "true"}
}

joined_3%>%
  subset(match %in% "true")%>%
  mutate(condition_comp = factor(condition_comp, 
                                 levels=c("feces", "7C",   "3C+Muc", "7C-Muc" , "3C" ,   "Glc"  ,"H2O"
                                          , "BHI like", "GMM like"
                                          )))-> beta_dist_c_sources
conidition_cond<-unique(beta_dist_c_sources$condition)

for (i in 1:8) {
beta_dist_c_sources%>%
  subset(condition %in% conidition_cond[i])->sub  
 print(shapiro_test(sub$value))
}


  

beta_dist_c_sources%>%
  t_test(data =., value ~ condition)%>%
  add_significance()%>%
  add_xy_position() %>%

###reordering of the significance labels
     mutate(xmin = ifelse(group1 == "7C" & group2 =="H2O", 1 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "7C" & group2 =="H2O",6, xmax))%>%
  mutate(y.position = ifelse(group1 == "7C" & group2 =="H2O",1, y.position))%>%
  
  mutate(xmin = ifelse(group1 == "3C+Muc" & group2 =="H2O", 2 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "3C+Muc" & group2 =="H2O",6, xmax))%>%
  mutate(y.position = ifelse(group1 == "3C+Muc" & group2 =="H2O",1.05, y.position))%>%
  
    mutate(xmin = ifelse(group1 == "7C-Muc" & group2 =="H2O", 3 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "7C-Muc" & group2 =="H2O",6, xmax))%>%
  mutate(y.position = ifelse(group1 == "7C-Muc" & group2 =="H2O", 1.1, y.position))%>%
  
   mutate(xmin = ifelse(group1 == "3C" & group2 =="H2O" , 4 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "3C" & group2 =="H2O",6, xmax))%>%
  mutate(y.position = ifelse(group1 == "3C" & group2 =="H2O",1.15, y.position))%>% 
  
    mutate(xmin = ifelse(group1 == "BHI like" & group2 =="H2O", 6 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "BHI like" & group2 =="H2O",7, xmax))%>%
  mutate(y.position = ifelse(group1 == "BHI like" & group2 =="H2O", 1.2, y.position)) %>%
  
  mutate(xmin = ifelse(group1 == "GMM like" & group2 =="H2O", 6 ,xmin))%>%
  mutate(xmax = ifelse(group1 == "GMM like" & group2 =="H2O",8, xmax))%>%
  mutate(y.position = ifelse(group1 == "GMM like" & group2 =="H2O", 1.25, y.position))->stat.test

beta_dist_c_sources%>%
  ggplot(aes(y=value, x=condition_comp))+
  geom_boxplot(aes(group=condition_comp),  outlier.shape =  NA)+
  geom_jitter(aes(group=condition_comp, colour=donor_name), width= 0.2, alpha=0.75, size = 3)+
  facet_grid(cols=vars(sample_type), space="free_x", scale="free_x")+
  theme(text=element_text(size=20),
        axis.title.x = element_blank(),
        axis.text.x = element_text( angle=90, vjust=0.2, hjust=1, size=20),
        strip.text.x = element_blank(),
        axis.ticks.x = element_blank(), 
        panel.background = element_rect(fill="white"),
        panel.border=element_rect(fill=NA),
        axis.title.y = element_text(size=20), 
        axis.text = element_text(size=20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size= 20),
        plot.title = element_text(size=20),
        strip.text.y = element_text(size=20),
        )+
  scale_x_discrete(labels = c('feces' = 'Feces', '7C' = '6C+Muc', '3C+Muc' = '3C+Muc', "Nut"="Nut", "SS"="SS", "7C-Muc"="6C", "3C"="3C", "H2O"="C-depleted", "Glc"="Glucose", "BHI like" = "BHI-like", "GMM like"= "GMM-like"))+
  ggtitle("Distances between feces and cultures")+
  ylab("Distance")+
  stat_pvalue_manual(stat.test, hide.ns = T, label = "p.adj.signif", size=5, label.size = 8)+
  labs(colour="Donor")->p

  show(p)
  
  ggsave(filename = "Beta_diversity_dist.png",
       path = "C:/Users/zuendj/Desktop/03_data R/wp1/output/final",
       width = 17,
       height = 17,
       units = c("cm"))
```

```{r}
aggregate(beta_dist_c_sources$value, by=list(beta_dist_c_sources$condition), FUN=median)%>%
  mutate(x = round(x, digits=2))
```

## Analysis shared taxa
## number of ASV culture vs. feces
```{r}
donor_count<-c("JHB", "KPJ", "WPL","SUT" ,"BFW" ,"TLM" ,"GCB", "AXZ")
maintain<-as.data.frame(matrix(5,0,0))

for (j in 1:length(donor_count)){
phyloseq_rare %>% 
  subset_samples(donor_name %in% c(donor_count[j]))%>%
    tax_glom(taxrank = "Genus")%>%
    transform_sample_counts(function(x) {x/sum(x)}) %>%
    filter_taxa(., function(x){mean(x) > 0}, TRUE)%>%
    psmelt()%>%
    mutate(condition = as.character(condition))%>%
    mutate(sample_type = as.character(sample_type))->df_abund
  


condition_count <- c("3C+Muc" ,  "7C-Muc" ,  "Glc"  ,    "3C"   ,    "feces"   , "H2O"    ,  "7C"   
                       , "GMM like" ,"BHI like"
                     )

abundance = 0.00
n=nrow(maintain)+1

for(i in 1:length(condition_count)){

  
df_abund%>%
   subset(sample_type %in% c("culture")) %>%
  subset(condition %in% c(condition_count[i]))%>%
  subset(Abundance > 0.00)-> culture

df_abund%>%
  subset(sample_type %in% c("feces"))%>%
  subset(Abundance > 0.00)-> feces

rbind(culture, feces)-> all

otu_all<-setNames(as.data.frame(unique(all$Genus)), c("Genus"))




diff_feces<-setNames(as.data.frame(setdiff(otu_all$Genus, culture$Genus)), c("Genus"))
diff_culture<-setNames(as.data.frame(setdiff(otu_all$Genus, feces$Genus)), c("Genus"))

intersec<-setNames(as.data.frame(intersect(culture$Genus, feces$Genus)), c("Genus"))

maintain[n,"donor_name"] = donor_count[j]
maintain[n,"condition"] = condition_count[i]

maintain[n,c("Shared")]<-nrow(intersec)/(nrow(diff_culture)+nrow(diff_feces)+nrow(intersec)) 
maintain[n,c("Cultures")]<-nrow(diff_culture)/(nrow(diff_culture)+nrow(diff_feces)+nrow(intersec)) 
maintain[n,c("Feces")]<-nrow(diff_feces)/(nrow(diff_culture)+nrow(diff_feces)+nrow(intersec)) 

n=n+1
  }}

```

```{r}
maintain%>%
  subset(condition %in% "7C")->C

mean(C$Shared)

left_join(setNames(aggregate(maintain$Shared, by = list(c(maintain$condition)), FUN="mean"), c("condition", "mean"))%>%
  mutate(mean = round(mean, digits=2)), 
  setNames(aggregate(maintain$Shared, by = list(c(maintain$condition)), FUN="sd"), c("condition", "sd"))%>%
  mutate(sd= round(sd, digits=2)) )
```

```{r}
setwd(mydir)

maintain%>%
  mutate(condition = factor(condition, levels = c("feces", "7C",   "3C+Muc", "7C-Muc" , "3C" ,   "Glc"  ,"H2O"
                                                   , "BHI like", "GMM like"
                                                  )))%>%
  mutate(Shared = Shared*100)%>%
  mutate(Cultures = Cultures*100)%>%
  mutate(Feces = Feces*100)%>%
  gather(sample, percent, c("Shared":"Feces"))%>%
  ggplot(aes(x=condition, y=percent, fill=sample))+
  geom_bar(stat = "identity", width=0.9143, colour="black", size=0.1 )+
  facet_grid(cols=vars(donor_name), scales="free_x", space = "free_x")+
  xlab("")+
  ylab("ASV representation[%]")+
  scale_x_discrete(labels = c('feces' = 'Feces', '7C' = '6C+Muc', '3C+Muc' = '3C+Muc', "Nut"="Nut", "SS"="SS", "7C-Muc"="6C", "3C"="3C", "H2O"="C-depleted", "Glc"="Glucose", 
                               "BHI like" = "BHI-like", "GMM like"= "GMM-like"
                              ))+
  scale_y_continuous(breaks=c(25, 50, 75))+
  scale_fill_manual(values=c("grey","burlywood4", "cyan4"))+
  theme(text=element_text(size=20),
        axis.text.x = element_text(size=17, angle=90, vjust=0.3, hjust=1),
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=20),
        axis.ticks.x = element_blank(),
        strip.text.x = element_blank(),
        axis.title.x = element_blank(), 
        panel.spacing = unit(0.05, "cm"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank())+
  labs(fill= "Sample")->p



    ggsave(filename = "maintained_asv.png",
       path = mydir,
       width = 40,
       height = 17,
       units = c("cm"))
   
p+labs(fill= "ASV representation")+ ggtitle("")+theme(legend.position = "none")



    ggsave(filename = "maintained_asv_side_plot_noleg..jpeg",
       path = "C:/Users/zuendj/Desktop/03_data R/wp1/output/final",
       width = 39.3,
       height = 10,
       units = c("cm"))


```
# Combine taxa bar plots & shared ASVs plot
```{r}
plot_grid(p_hist           + theme(axis.text.x = element_blank(),
                        axis.ticks.x  = element_blank())
          , 
          p          + theme(strip.text = element_blank())
         ,
  align= "v", rel_heights = c(0.6,0.4), rel_widths=c(1,1), nrow=2, axis = 'tb')

 ggsave(filename = "C_sources_overall.jpeg",
       path = "C:/Users/zuendj/Desktop/03_data R/wp1/output/final",
       width = 50,
       height = 24,
       units = c("cm"))

```




